---
title: "final_project"
author: "brett"
date: "April 9, 2019"
output: html_document
---

Brett Waugh
9 April 2019
adult_logistic.R
Looks at data about people in the United States in 1996 and makes a prediction about income.
Most of script is taken from https://rpubs.com/H_Zhu/235617 with very few revisions. 

```{r, include=FALSE}
# Load necessary libraries. 
require(ggplot2)
require(plyr)
require(ROCR)
require(dplyr)
require(ISLR)
require(boot)
require(caret)
```

Get the data from here: https://rpubs.com/H_Zhu/235617. Then read in the data. 

```{r, include=FALSE}
# Get the data from the source.
adult <- read.table('https://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data', 
                    sep = ',', fill = F, strip.white = T)
colnames(adult) <- c('age', 'workclass', 'fnlwgt', 'educatoin', 
                     'educatoin_num', 'marital_status', 'occupation', 'relationship', 'race', 'sex', 
                     'capital_gain', 'capital_loss', 'hours_per_week', 'native_country', 'income')
```

After adding in the data, check it to make sure everything looks correct. 
```{r, include=TRUE}
str(adult)
```

Next, we will remove some variables and create some graphs. 
```{r, include=TRUE}
### Three variables dropped
adult$educatoin <- NULL
adult$fnlwgt <- NULL
adult$relationship <- NULL
```

The first graph is a histogram of age by income group. 
```{r, include=TRUE}
# Histogram of age by income group. 
ggplot(adult) + aes(x=as.numeric(age), group=income, fill=income) + 
  geom_histogram(binwidth=1, color='black')
```
As you can see in this graph, there are drastically more people making under 50k than above 50k.
The center for those making below 50k is around 32, while the center for those making above 50k is around 34. 
This is a small difference in age, which may mean there are other variables in play here.

The next graph is a histogram of age by gender group. 
```{r, include=TRUE}
# Histogram of age by gender group.
ggplot(adult) + aes(x=as.numeric(age), group=sex, fill=sex) + 
  geom_histogram(binwidth=1, color='black')
```
This graph immediately shows that there are more female participants than male participants in the study.
This could be representative of the overall population trend, but it may also skew the results of the study. 
There does not appear to be a significant difference with regard to age between the two sexes. 

```{r, include=FALSE}
summary(adult$workclass)
levels(adult$workclass)[1] <- 'Unknown'

# Combine into Government job.
adult$workclass <- gsub('^Federal-gov', 'Government', adult$workclass)
adult$workclass <- gsub('^Local-gov', 'Government', adult$workclass)
adult$workclass <- gsub('^State-gov', 'Government', adult$workclass) 

# Combine into Sele-Employed job.
adult$workclass <- gsub('^Self-emp-inc', 'Self-Employed', adult$workclass)
adult$workclass <- gsub('^Self-emp-not-inc', 'Self-Employed', adult$workclass)

# Combine into Other/Unknown.
adult$workclass <- gsub('^Never-worked', 'Other', adult$workclass)
adult$workclass <- gsub('^Without-pay', 'Other', adult$workclass)
adult$workclass <- gsub('^Other', 'Other/Unknown', adult$workclass)
adult$workclass <- gsub('^Unknown', 'Other/Unknown', adult$workclass)

adult$workclass <- as.factor(adult$workclass)

summary(adult$workclass)

# Get the counts by industry and income group
count <- table(adult[adult$workclass == 'Government',]$income)["<=50K"]
count <- c(count, table(adult[adult$workclass == 'Government',]$income)[">50K"])
count <- c(count, table(adult[adult$workclass == 'Other/Unknown',]$income)["<=50K"])
count <- c(count, table(adult[adult$workclass == 'Other/Unknown',]$income)[">50K"])
count <- c(count, table(adult[adult$workclass == 'Private',]$income)["<=50K"])
count <- c(count, table(adult[adult$workclass == 'Private',]$income)[">50K"])
count <- c(count, table(adult[adult$workclass == 'Self-Employed',]$income)["<=50K"])
count <- c(count, table(adult[adult$workclass == 'Self-Employed',]$income)[">50K"])
count <- as.numeric(count)

# Create a dataframe.
industry <- rep(levels(adult$workclass), each = 2)
income <- rep(c('<=50K', '>50K'), 4)
df <- data.frame(industry, income, count)
df

# Calculate the percentages.
df <- ddply(df, .(industry), transform, percent = count/sum(count) * 100)

# Format the labels and calculate their positions.
df <- ddply(df, .(industry), transform, pos = (cumsum(count) - 0.5 * count))
df$label <- paste0(sprintf("%.0f", df$percent), "%")
```


After preparing the data more, and combining several of the fields we can create a barplot for the data. 
```{r, include=TRUE}
# Barplot of counts by industry with in group proportions.
ggplot(df, aes(x = industry, y = count, fill = income)) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = pos, label = label), size = 2) + 
  ggtitle('Income by Industry')
```
From this barplot, we can see that the Private industry employees the most workers, followed by the government, then self-employed, then other/unknown.
The Private industry also has the greatest number of people making above 50k per year, but Self-Employed industries have the greatest percentage of people making above 50k. 
The industry with the most number of people making 50k or less is Other/Unknown.

```{r, include=FALSE}
# Create a dataframe.
df1 <- data.frame(table(adult$income, adult$educatoin_num))
names(df1) <- c('income', 'education_num', 'count')
df1

# Calculate the percentages.
df1 <- ddply(df1, .(education_num), transform, percent = count/sum(count) * 100)

# Format the labels and calculate their positions.
df1 <- ddply(df1, .(education_num), transform, pos = (cumsum(count) - 0.5 * count))
df1$label <- paste0(sprintf("%.0f", df1$percent), "%")

# Remove some in group percentage to avoid overlapped text.
df1$label[which(df1$percent < 5)] <- NA
```

After preparing an additional dataframe and calculating some values, another barplot can be created. 
```{r, include=TRUE}
# Barplot of counts by years of education with in group proportions.
ggplot(df1, aes(x = education_num, y = count, fill = income)) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = pos, label = label), size = 2) + 
  ggtitle('Income Level with Years of Education')
```
Immediately, it is apparent that the greatest number of people making 50k or less have nine or ten years of education. 
There are few individuals in the study with less than nine years of education making above 50k. 
The number of years with the greatest percentage above 50k are sixteen years, with fifteen years coming in a close second. 
From this study, 100% of those with only one year of education make 50k or less. 

```{r, include=FALSE}
summary(adult$occupation)

levels(adult$occupation)[1] <- 'Unknown'
adult$occupation <- gsub('Adm-clerical', 'White-Collar', adult$occupation)
adult$occupation <- gsub('Craft-repair', 'Blue-Collar', adult$occupation)
adult$occupation <- gsub('Exec-managerial', 'White-Collar', adult$occupation)
adult$occupation <- gsub('Farming-fishing', 'Blue-Collar', adult$occupation)
adult$occupation <- gsub('Handlers-cleaners', 'Blue-Collar', adult$occupation)
adult$occupation <- gsub('Machine-op-inspct', 'Blue-Collar', adult$occupation)
adult$occupation <- gsub('Other-service', 'Service', adult$occupation)
adult$occupation <- gsub('Priv-house-serv', 'Service', adult$occupation)
adult$occupation <- gsub('Prof-specialty', 'Professional', adult$occupation)
adult$occupation <- gsub('Protective-serv', 'Service', adult$occupation)
adult$occupation <- gsub('Tech-support', 'Service', adult$occupation)
adult$occupation <- gsub('Transport-moving', 'Blue-Collar', adult$occupation)
adult$occupation <- gsub('Unknown', 'Other/Unknown', adult$occupation)
adult$occupation <- gsub('Armed-Forces', 'Other/Unknown', adult$occupation)
adult$occupation <- as.factor(adult$occupation)
summary(adult$occupation)

# Create a dataframe.
df2 <- data.frame(table(adult$income, adult$occupation))
names(df2) <- c('income', 'occupation', 'count')
df2

# Calculate the percentages.
df2 <- ddply(df2, .(occupation), transform, percent = count/sum(count) * 100)

# Format the labels and calculate their positions.
df2 <- ddply(df2, .(occupation), transform, pos = (cumsum(count) - 0.5 * count))
df2$label <- paste0(sprintf("%.0f", df2$percent), "%")
```

```{r, include=TRUE}
# Barplot of counts by occupation with in group proportions.
ggplot(df2, aes(x = occupation, y = count, fill = income)) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = pos, label = label), size = 2) + 
  ggtitle('Income Level with Different Occupations')
```
Income level can very greatly depending on which field someone goes into. 
From this dataset, the greatest number of people making 50k or less is from the Blue-Collar occupations. 
The greatest number of people making more than 50k are in White-Collar occupations. 
While the greatest number may be in White-Collar, Professional occupations holds the greatest percentage of people making above 50k. 
The occupations with the greatest percentage making 50k or less is in Other/Unknown and Service. 

```{r, include=FALSE}
summary(adult$marital_status)

adult$marital_status <- gsub('Married-AF-spouse', 'Married', adult$marital_status)
adult$marital_status <- gsub('Married-civ-spouse', 'Married', adult$marital_status)
adult$marital_status <- gsub('Married-spouse-absent', 'Married', adult$marital_status)
adult$marital_status <- gsub('Never-married', 'Single', adult$marital_status)
adult$marital_status <- as.factor(adult$marital_status)
summary(adult$marital_status)

df3 <- data.frame(table(adult$income, adult$marital_status))
names(df3) <- c('income', 'marital_status', 'count')
df3

# Calculate the percentages.
df3 <- ddply(df3, .(marital_status), transform, percent = count/sum(count) * 100)

# Format the labels and calculate their positions.
df3 <- ddply(df3, .(marital_status), transform, pos = (cumsum(count) - 0.5 * count))
df3$label <- paste0(sprintf("%.0f", df3$percent), "%")
```

```{r, include=TRUE}
# Barplot of counts by marital status with in group proportions.
ggplot(df3, aes(x = marital_status, y = count, fill = income)) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = pos, label = label), size = 2) + 
  ggtitle('Income Level with Marital Status')
```
Marital status appears to play a significant role in determing how much someone will make.
The greatest number of people making both types of income are Married. 
Married also has the greatest percentage of individuals making more than 50k. 
The greatest percentage of people making less than 50k is Single. However, Divorced, Seperated, Single, and Widowed also all have above 90% making 50k or less. 

Looking at the histograms for capital gains and capital losses, it is clear that very few people partake in this. 
Due to how few individuals partake in it, it is best to remove it from the study to not skew the results. 
```{r, include=TRUE}
# Histogram of capital_gain.
ggplot(adult) + aes(x=as.numeric(capital_gain), group=income, fill=income) + 
  geom_histogram(bins=10, color='black') + ggtitle('Histogram of Capital Gain')

# Histogram of capital_loss.
ggplot(adult) + aes(x=as.numeric(capital_loss), group=income, fill=income) + 
  geom_histogram(bins=10, color='black') + ggtitle('Histogram of Capital Loss')

# Percentage of observatiosn with no capital gain or loss.
sum(adult$capital_gain == 0)/length(adult$capital_gain)

sum(adult$capital_loss == 0)/length(adult$capital_loss)
```

Native country will also be removed. 

```{r, include=FALSE}
# Variables appear to be unimportant for analysis and are dropped. 
adult$capital_gain <- NULL
adult$capital_loss <- NULL
adult$native_country <- NULL

df4 <- data.frame(table(adult$income, adult$race))
names(df4) <- c('income', 'race', 'count')
df4

# Calculate the percentages.
df4 <- ddply(df4, .(race), transform, percent = count/sum(count) * 100)

# Format the labels and calculate their positions.
df4 <- ddply(df4, .(race), transform, pos = (cumsum(count) - 0.5 * count))
df4$label <- paste0(sprintf("%.0f", df4$percent), "%")

# Do not display percentage for low counts categories.
df4$label[df4$race == 'Other'] <- NA
df4$label[df4$race == 'Amer-Indian-Eskimo'] <- NA
```

```{r, include=TRUE}
# Bar plot of counts by marital status with in group proportions.
ggplot(df4, aes(x = race, y = count, fill = income)) +
  geom_bar(stat = "identity") +
  geom_text(aes(y = pos, label = label), size = 2) + 
  ggtitle('Income Level by Race')
```
Looking at income distribution by race can be a tricky situation as some people can be from multple races, or are sometimes unaware of ancestry. 
The greatest number of people making above 50k are White, but the greatest number of people making 50k or less are also White. 
Amer-Indian-Eskimo, Asian-Pac-Islander, Other, and Black all have high percentages of people making 50k or less. 

At this point, we only have nine remaining features to work with.
```{r, include=TRUE}
summary(adult)
```

The next stage in the process is creating a model. For this particular dataset, logistic regression will be used. 
```{r, include=FALSE}
# 80% of the original data is used as the training set, while the remaining 20% is used as test set. 
sz <- round(.8 * dim(adult)[1])  # training set size
training_set <- adult[1:sz,]
testing_set <- adult[-(1:sz),]

# Income will be the response variable, the other eight variables will be fit.
m1 <- glm(income ~ ., data = training_set, family = binomial('logit'))
```
```{r, include=TRUE}
summary(m1)
confint(m1)
```
We can see the relationships between some of the variables here.
Notice some of the features with the low P Values: age, education_num, martital_StatusMarried, martitalStatusSingle, occupationProfessional, OccupationSales, occupationWhite-Collar, sexMale, and hours_per_week. 
These features seem to have a significant impact on determing the income of an individual. 

```{r, include=FALSE}
# Full model is the model just fitted.
m_full <- m1  
m_null <- glm(income ~ 1, data = training_set, family = binomial('logit'))

# Backward selection.
step(m_full, trace = F, scope = list(lower=formula(m_null), upper=formula(m_full)),
     direction = 'backward')

# Forward selection.
step(m_null, trace = F, scope = list(lower=formula(m_null), upper=formula(m_full)),
     direction = 'forward')

#---- Logistic model ---- #
prob <- predict(m1, testing_set, type = 'response')
pred <- rep('<=50K', length(prob))
pred[prob>=.5] <- '>50K'
```
```{r, include=TRUE}
# confusion matrix 
tb <- table(pred, testing_set$income)
tb
(4533+855)/(4533+745+379+855)
```

From the confusion matrix we can determine the accuracy of the model to be 0.827 or 82.7%.
This is a great score for the model, because it does not suggest overfitting while maintaining a high degree of accuracy. 
Given information about an individual, the model should be able to accurately predict if the individual makes greater than 50k 82.7% of the time.

The second model uses a slightly different approach to the problem. 
```{r, include=FALSE}
# ---- Second model ---- #
ControlParameters <-trainControl(method="repeatedcv", 
                                 number=10,
                                 repeats=10)

otherModel <-train(income ~ ., 
                data=training_set,
                method="glm",
                trControl= ControlParameters
)

```
```{r, include=TRUE}
otherModel
```
While this model took a different approach to the problem, it still yielded an accuracy of 0.824 or 82.4%. 
Another benefit to this model is that it uses 10-fold cross validation, making it understandable to lose an additional percentage from the first model.